Implementation of Parallel Algorithms for ARIMA in Distributed Database Systems

### 2013-09-24 draft.tex
Made it compilable
Updates

\documentclass[english,12pt]{article}
\usepackage{babel}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{algorithm2e}

@@

At the time of writing, MADlib has modules for linear regression, logistic
regression, multinomial logistic regression, elastic-net regularization for
linear and logistic regressions, all kinds of robust estimators for
regressions, marginal effects, k-means, association rules, cross validation,
linear systems, matrix factorization, LDA, data summary, correlation,
hypothesis tests, SVD and PCA, ARIMA etc and many other supporting functions.
A very detailed user documentation is available onlin at
\url{http://madlib.net}.

@@

@@

The likelihood function $L$ for $N$ values of $Z_t$  is then
\begin{equation}
L(\phi, \theta) = \prod_{t = 1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}
\end{equation}
so the log likelihood function $l$ is
\begin{eqnarray}
l(\phi, \theta) &=& \sum_{t = 1}^N \ln \left(\frac{1}{\sqrt{2 \pi \sigma^2}}
 e^{-Z^2_t/2 \sigma^2}
 \right) \nonumber\\
 &=&  \sum_{t = 1}^N  - \ln \left( \sqrt{2 \pi \sigma^2}\right)
-\frac{Z^2_t}{2 \sigma^2}\nonumber\\
&=&  -\frac{N}{2} \ln \left( 2 \pi \sigma^2\right)  - \frac{1}{2
  \sigma^2} \sum_{t = 1}^N   Z^2_t\ . \label{eq:loglikelihood}
\end{eqnarray}
Thus, finding the maximum likelihood is equivalent to solving the
optimization problem (known as the conditional least squares
formation)
\begin{equation}
\min_{\theta, \phi} \sum_{t = 1}^N  Z^2_t.
\end{equation}
The error term $Z_t$ can be computed iteratively as follows:
\begin{equation}
    Z_t = X_t - F_t(\phi, \theta, \mu) \label{eq:error-terms}
\end{equation}
where
\begin{equation}
F_t(\phi, \theta, \mu) = \mu + \sum_{i=1}^p \phi_i (X_{t-i}-\mu) + \sum_{i=1}^q
\theta_i Z_{t-i}
\end{equation}

@@

Eqs. (\ref{eq:error-terms},~\ref{eq:J1},~\ref{eq:J2},~\ref{eq:J3}) are
difficult to parallelize.

>>>>

\documentclass[english,12pt]{article}
\usepackage{babel}
\usepackage{natbib}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm2e}

@@

At the time of writing, MADlib has modules for linear regression, logistic
regression, multinomial logistic regression, elastic-net regularization for
linear and logistic regressions, all kinds of robust estimators for
regressions, marginal effects, k-means, association rules, cross validation,
linear systems, matrix factorization, LDA, data summary, correlation,
hypothesis tests, SVD and PCA, ARIMA etc and many other supporting functions.
A very detailed user documentation is available onlin at
http://madlib.net.

@@

In the next few subsections, we will first describe the algorithm that we have
used to solve the problem of maximization of partial log-likelihood. Then, we
point out the reason why it is difficult to make the algorithm run parallel.
The difficulty is not restricted to this specific algorithm. It is a
common situation in all algorithms that fit ARIMA models. Actually it is easy
to see that many machine learning models have similar problems. Next, we
describe our solution to this problem. Our method is quite general and can
parallelize similar problems. Then we describe a simple method to improve the
performance of our algorithm. In the end, we discuss the possible ways to
generalize our methods to more algorithms.

@@

The likelihood function $L$ for $N$ values of $Z_t$  is then
\begin{equation}
L(\phi, \theta) = \prod_{t = 1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}
\end{equation}
so the log likelihood function $l$ is
\begin{eqnarray}
l(\phi, \theta) &=& \sum_{t = 1}^N \ln \left(\frac{1}{\sqrt{2 \pi \sigma^2}}
 e^{-Z^2_t/2 \sigma^2}
 \right) \nonumber\\
 &=&  \sum_{t = 1}^N  - \ln \left( \sqrt{2 \pi \sigma^2}\right)
-\frac{Z^2_t}{2 \sigma^2}\nonumber\\
&=&  -\frac{N}{2} \ln \left( 2 \pi \sigma^2\right)  - \frac{1}{2
  \sigma^2} \sum_{t = 1}^N   Z^2_t\ \. \label{eq:loglikelihood}
\end{eqnarray}
Thus, finding the maximum likelihood is equivalent to solving the
optimization problem (known as the conditional least squares
formation)
\begin{equation}
\min_{\theta, \phi} \sum_{t = 1}^N  Z^2_t.
\end{equation}
The error term $Z_t$ can be computed iteratively as follows:
\begin{equation}
    Z_t = X_t - F_t(\phi, \theta, \mu) \label{eq:error-terms}
\end{equation}
where
\begin{equation}
F_t(\phi, \theta, \mu) = \mu + \sum_{i=1}^p \phi_i (X_{t-i}-\mu) + \sum_{i=1}^q
\theta_i Z_{t-i}
\end{equation}

@@

It is very easy to see that Eqs.
(\ref{eq:error-terms},~\ref{eq:J1},~\ref{eq:J2},~\ref{eq:J3}) are
difficult to parallelize. Each step computation uses the result from the
previous step. Therefore, we have to scan through the data sequentially to
compute the quatities in these equations.
